---
title: "Stack datasets"
subtitle: "bind_rows()"
---

::: {.sas-equiv}
**SAS:** `SET` with several datasets; `INDSNAME=`; `PROC SQL` `UNION` / `OUTER UNION CORR`
:::

`dplyr::bind_rows()` stacks tables vertically. Columns are matched by name.
Columns present in only one table are filled with `NA` in the other rows.

## Quick reference

{{< include _quickref/bind-rows.md >}}

Worked examples below.

## Stack two tables

```{r}
#| label: stack
library(dplyr)

ae_site_a <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002"),
  AETERM = c("NAUSEA", "HEADACHE"),
  AESEV = c("MILD", "MODERATE")
)

ae_site_b <- data.frame(
  USUBJID = c("STD-001-0003"),
  AETERM = c("FATIGUE"),
  AESEV = c("MILD"),
  AEREL = c("RELATED")
)

bind_rows(ae_site_a, ae_site_b)
```

`AEREL` is `NA` for site A rows. SAS `set ae_site_a ae_site_b;` is the usual
analogue when variable names already match.

## Mark the source table

```{r}
#| label: id
bind_rows(
  list(SITEA = ae_site_a, SITEB = ae_site_b),
  .id = "SITE"
)
```

`.id` adds a column with the list names. SAS:

```sas
data all;
  set ae_site_a ae_site_b indsname=src;
  site = src;
run;
```

`INDSNAME=` is the dataset name that contributed the row (often `WORK.AE_SITE_A`
style). Rename or `scan` that value if you need a short code like `.id`.

::: {.callout-important}
## Match by name, not position

`bind_rows()` does not line up column 1 with column 1. Rename first when the
same concept has different names. Character versus numeric on the same name
is an error in dplyr. Fix the type before stacking.
:::

## SAS truncates; `bind_rows()` does not

The first SAS dataset in `SET` sets character lengths. A later dataset with a
longer value is truncated to that length. `bind_rows()` keeps the full string:

```{r}
#| label: lengths
ae_short <- data.frame(
  USUBJID = "STD-001-0001",
  AETERM = "NAUSEA"
)

ae_long <- data.frame(
  USUBJID = "STD-001-0003",
  AETERM = "MYOCARDIAL INFARCTION"
)

bind_rows(ae_short, ae_long)
```

`AETERM` is the full term on both rows. In SAS, `set ae_short ae_long;` would
cut `MYOCARDIAL INFARCTION` to the length of `NAUSEA` unless you set length
first.

For side-by-side combines on keys, use joins (see Joins), not `bind_rows()`.
