---
title: "Bind rows"
subtitle: "Stack tables on top of each other"
---

::: {.sas-equiv}
**SAS:** `SET` with several datasets; `PROC SQL` `UNION` / `OUTER UNION CORR`
:::

`dplyr::bind_rows()` stacks tables vertically. Columns are matched by name.
Columns present in only one table are filled with `NA` in the other rows.

## Quick reference

{{< include _quickref/bind-rows.md >}}

Worked examples below.

## Stack two tables

```{r}
#| label: stack
library(dplyr)

ae_site_a <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002"),
  AETERM = c("NAUSEA", "HEADACHE"),
  AESEV = c("MILD", "MODERATE")
)

ae_site_b <- data.frame(
  USUBJID = c("STD-001-0003"),
  AETERM = c("FATIGUE"),
  AESEV = c("MILD"),
  AEREL = c("RELATED")
)

bind_rows(ae_site_a, ae_site_b)
```

`AEREL` is `NA` for site A rows. SAS `set ae_site_a ae_site_b;` is the usual
analogue when variable names already match.

## Mark the source table

```{r}
#| label: id
bind_rows(
  list(SITEA = ae_site_a, SITEB = ae_site_b),
  .id = "SITE"
)
```

`.id` adds a column with the list names. Useful when you need to know which
input row came from.

::: {.callout-important}
## Match by name, not position

`bind_rows()` does not line up column 1 with column 1. Rename first when the
same concept has different names. When the same name has different types,
fix the type before stacking.
:::

For side-by-side combines on keys, use joins (see Joins), not `bind_rows()`.
