---
title: "Coalesce"
subtitle: "Prefer one value, fall back to another"
---

::: {.sas-equiv}
**SAS:** `COALESCE`, `coalescec`, `IFN` / `IFC`, specify-then-coded
:::

SAS `COALESCE` / `coalescec` skip missing values. `dplyr::coalesce()` skips `NA`
only. Blank `""` is a value, not missing.

## Quick reference

{{< include _quickref/coalesce.md >}}

Worked examples below.

## The example table

Specify versus coded location. Four rows; four states of the specify field.

```{r}
#| label: setup
library(dplyr)

cm <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002", "STD-001-0003", "STD-001-0004"),
  LOCSPY  = c("Left deltoid", "", NA, "Right thigh, anterior"),
  LOC     = c("ARM", "ARM", "LEG", "THIGH")
)

cm
```

- `STD-001-0001`: specify present; overrides `ARM`
- `STD-001-0002`: specify is `""`; use coded
- `STD-001-0003`: specify is `NA`; use coded
- `STD-001-0004`: specify present; overrides `THIGH`

## `coalesce()` skips `NA`

```sas
loc = coalescec(locspy, loc);
```

```{r}
cm |>
  mutate(LOC_OUT = coalesce(LOCSPY, LOC))
```

`STD-001-0003` is correct: `NA` is skipped, `LEG` is used. `STD-001-0002` is not.
`""` is not `NA`, so `coalesce()` keeps it and does not use coded `ARM`.

::: {.callout-important}
## `coalesce()` does not treat `""` as missing

SAS character missing is blank. R character missing is `NA`. `coalesce()` follows R.
A specify field of `""` is present, so it overrides the coded term. Convert blanks
with `na_if()`, then coalesce.
:::

## The fix: `na_if()` then `coalesce()`

```{r}
cm |>
  mutate(LOC_OUT = coalesce(na_if(LOCSPY, ""), LOC))
```

Specify is used when it has text. Coded is used when specify is `""` or `NA`.

Specify-overrides-coded is `coalesce(na_if(spy, ""), coded)`. Reversed arguments
`coalesce(coded, na_if(spy, ""))` keep `ARM` on `STD-001-0001` and drop the specify
text.

## `na_if()` on its own

```{r}
na_if(cm$LOCSPY, "")
```

`""` becomes `NA`. Other strings are unchanged. `NA` remains `NA`.

The characters `"NA"` are not missing. If an export stored that string, convert it
with `na_if(x, "NA")`. That is separate from `na_if(x, "")`.

## `IFN` / `IFC`

SAS `IFN` (numeric) and `IFC` (character) are a condition and two (or three) results.
That mapping is `if_else()`, not `coalesce()`:

```sas
loc = ifc(missing(locspy) or locspy = "", loc, locspy);
```

```{r}
cm |>
  mutate(
    LOC_OUT = if_else(
      is.na(na_if(LOCSPY, "")),
      LOC,
      LOCSPY
    )
  )
```

Same result as `coalesce(na_if(LOCSPY, ""), LOC)`. Use `coalesce()` for first
non-missing. Use `if_else()` when the rule is a named condition.
