---
title: "Coalesce"
subtitle: "Prefer one value, fall back to another"
---

::: {.sas-equiv}
**SAS:** `COALESCE`, `coalescec`, `IFN` / `IFC`, specify-then-coded
:::

SAS `COALESCE` / `coalescec` skip missing values. R's `dplyr::coalesce()` skips `NA`
only. Blank `""` is a real value, so it wins. That is the whole page.

## Quick reference

{{< include _quickref/coalesce.md >}}

The rest of this page is the same material worked through slowly, with runnable examples
and the output they produce. Skip it if the block above was what you came for.

## The example table

Specify versus coded location. Four rows, four states of the specify field.

```{r}
#| label: setup
library(dplyr)

cm <- data.frame(
  USUBJID = c("XYZ-101-0001", "XYZ-101-0002", "XYZ-101-0003", "XYZ-101-0004"),
  LOCSPY  = c("Left deltoid", "", NA, "Right thigh, anterior"),
  LOC     = c("ARM", "ARM", "LEG", "THIGH")
)

cm
```

- `0001` has a specify value, which should override `ARM`
- `0002` has `""` on the specify field: use the coded value
- `0003` has `NA` on the specify field: also use the coded value
- `0004` has a specify value, which should override `THIGH`

## `coalesce()` skips `NA`

```sas
loc = coalescec(locspy, loc);
```

```{r}
cm %>%
  mutate(LOC_OUT = coalesce(LOCSPY, LOC))
```

`0003` is correct: `NA` is skipped, `LEG` comes through. `0002` is not. `""` is not `NA`,
so `coalesce()` keeps it and the coded `ARM` never gets a look.

::: {.callout-important}
## `coalesce()` does not treat `""` as missing

SAS character missing is blank. R character missing is `NA`. `coalesce()` follows R.
A specify field of `""` is a present value, so it beats the coded term. Convert blanks
first with `na_if()`, then coalesce.
:::

## The fix: `na_if()` then `coalesce()`

```{r}
cm %>%
  mutate(LOC_OUT = coalesce(na_if(LOCSPY, ""), LOC))
```

Four rows, four coded-or-specify results. Specify wins when it has text. Coded wins
when specify is `""` or `NA`.

That is the specify-overrides-coded pattern: `coalesce(na_if(spy, ""), coded)`. Do not
reverse the arguments. `coalesce(coded, na_if(spy, ""))` would keep `ARM` on `0001` and
throw the free text away.

## `na_if()` on its own

```{r}
na_if(cm$LOCSPY, "")
```

`""` becomes `NA`. Everything else, including a real specify string, is untouched.
`NA` in is `NA` out.

The literal characters `"NA"` are also not missing. If a CRF export wrote that string,
`na_if(x, "NA")` is a separate conversion, not a side effect of `na_if(x, "")`.

## `IFN` / `IFC`

SAS `IFN` (numeric) and `IFC` (character) are a condition plus two (or three) results.
That is `if_else()`, not `coalesce()`:

```sas
loc = ifc(missing(locspy) or locspy = "", loc, locspy);
```

```{r}
cm %>%
  mutate(
    LOC_OUT = if_else(
      is.na(na_if(LOCSPY, "")),
      LOC,
      LOCSPY
    )
  )
```

Same output as `coalesce(na_if(LOCSPY, ""), LOC)`. Prefer `coalesce()` when the rule
is "first non-missing", and `if_else()` when the rule is a condition you can name.
