---
title: "Filter"
subtitle: "Keep rows that match a condition"
---

::: {.sas-equiv}
**SAS:** `WHERE`, subsetting `IF`
:::

`dplyr::filter()` keeps rows for which the condition is `TRUE`.

## Quick reference

{{< include _quickref/filter.md >}}

Worked examples below.

## One condition

```{r}
#| label: one
library(dplyr)

vs <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0001", "STD-001-0002"),
  PARAMCD = c("SYSBP", "DIABP", "SYSBP"),
  AVAL = c(120, 80, NA_real_)
)

vs |>
  filter(PARAMCD == "SYSBP")
```

SAS: `where paramcd = "SYSBP";`

## Several conditions

```{r}
#| label: several
vs |>
  filter(PARAMCD == "SYSBP", !is.na(AVAL))
```

Commas in `filter()` mean AND. Use `|` for OR. `!is.na(AVAL)` is the usual way
to drop missing results before a comparison.

## Missing numeric: SAS `WHERE` keeps the row

Same table. Compare `AVAL < 100` in both languages.

```{r}
#| label: missing
vs |>
  filter(AVAL < 100)
```

| `AVAL` | SAS `where aval < 100` | R `filter(AVAL < 100)` |
|---|---|---|
| 120 | drop | drop |
| 80 | keep | keep |
| missing | keep | drop |

R keeps only 80. `NA < 100` is `NA`, not `TRUE`, so `filter()` drops that row.

SAS `where aval < 100;` keeps 80 and the missing row. A missing numeric is
smaller than any number, so the comparison is true.

To drop missing in R, say so:

```{r}
#| label: drop-na
vs |>
  filter(!is.na(AVAL), AVAL < 100)
```

To keep missing on purpose (the SAS `WHERE` result on this data):

```{r}
#| label: keep-na
vs |>
  filter(is.na(AVAL) | AVAL < 100)
```

::: {.callout-important}
## `NA` in the condition drops the row

`filter(AVAL < 100)` drops rows where `AVAL` is `NA`. SAS `WHERE aval < 100`
keeps them. Write `!is.na(AVAL)` when missing must not pass.
:::
