---
title: "If else and case when"
subtitle: "Conditional column values"
---

::: {.sas-equiv}
**SAS:** `IF-THEN-ELSE`, `IFN` / `IFC`, `SELECT`
:::

Use `if_else()` for one condition and two results. Use `case_when()` for several
branches.

## Quick reference

{{< include _quickref/if-else-case-when.md >}}

Worked examples below.

## Two-way: `if_else()`

```{r}
#| label: ifelse
library(dplyr)

dm <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002", "STD-001-0003"),
  AGE = c(54L, 70L, NA_integer_)
)

dm |>
  mutate(AGEGR1 = if_else(AGE < 65L, "<65", ">=65"))
```

When `AGE` is `NA`, `if_else()` returns `NA` for that row (unless you set
`missing=`).

SAS `ifc(age < 65, "<65", ">=65")` is the same idea for character results.

## Multi-way: `case_when()`

```{r}
#| label: casewhen
ae <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002", "STD-001-0003", "STD-001-0004"),
  AESEV = c("MILD", "MODERATE", "SEVERE", NA_character_)
)

ae |>
  mutate(
    AESEVN = case_when(
      AESEV == "MILD" ~ 1L,
      AESEV == "MODERATE" ~ 2L,
      AESEV == "SEVERE" ~ 3L,
      TRUE ~ NA_integer_
    )
  )
```

The first matching condition wins. End with `TRUE ~ ...` as the catch-all
(SAS OTHERWISE).

::: {.callout-important}
## Prefer `if_else()` over base `ifelse()` in dplyr

`if_else()` checks that true and false values share a type. Base `ifelse()` is
looser and often surprises with factors and dates.
:::

For "first non-missing of several columns," use `coalesce()` (see Coalesce), not
`case_when()`.
