---
title: "Joins"
subtitle: "Combining tables"
---

::: {.sas-equiv}
**SAS:** `DATA` step `MERGE ... BY`, `PROC SQL` `INNER`/`LEFT`/`RIGHT`/`FULL JOIN`,
`IN=` flags, `WHERE EXISTS`
:::

R has six named join functions where SAS has one `MERGE` statement plus a set of
conventions. Once you know which name maps to which `IN=` pattern, joins get shorter and
much harder to get silently wrong.

## The example tables

Two small tables with deliberate mismatches, used throughout this page.

```{r}
#| label: setup
library(dplyr)

dm <- data.frame(
  subjid = c("001", "002", "003", "004", "005"),
  sex    = c("M", "F", "F", "M", "M"),
  trt    = c("Placebo", "Active", "Active", "Placebo", "Active")
)

ae <- data.frame(
  subjid = c("001", "001", "002", "003", "006"),
  aeterm = c("Headache", "Nausea", "Rash", "Fatigue", "Dizziness"),
  sev    = c("Mild", "Moderate", "Severe", "Mild", "Mild")
)

dm
ae
```

The mismatches matter, because they are what the different joins disagree about:

- subject `001` has two AEs, so this is one-to-many
- subjects `004` and `005` are in `dm` with no AEs
- subject `006` has an AE but is **not** in `dm`, a data error you want to catch

## The six joins

### `inner_join()`: rows matching in both

```sas
proc sql;
  select * from dm inner join ae on dm.subjid = ae.subjid;
quit;

/* or, DATA step */
data out;
  merge dm(in=a) ae(in=b);
  by subjid;
  if a and b;
run;
```

```{r}
inner_join(dm, ae, by = "subjid")
```

Four rows. Dropped: `004` and `005` (no AE), and `006` (not in `dm`). Note that the data
error and the legitimately event-free subjects vanish together and without comment.
That is exactly why `inner_join()` makes a poor default.

### `left_join()`: keep everything on the left

```sas
data out;
  merge dm(in=a) ae;
  by subjid;
  if a;
run;
```

```{r}
left_join(dm, ae, by = "subjid")
```

This is the workhorse, because most real joins are "start from my subject list and
annotate it".
Six rows: `001` twice because it has two AEs, and `004`/`005` once each with `NA` in the
AE columns.

::: {.callout-important}
## `NA` from a join means "no match", not "missing value"

R gives you the same `NA` for both, and only context tells you which you have. Here it
means the subject had no adverse event. It does not mean the term was left blank on the CRF, and it does
not that the AE extract failed to load. If that distinction matters downstream, record it
explicitly with a flag column rather than relying on the `NA`.
:::

### `right_join()`: keep everything on the right

```{r}
right_join(dm, ae, by = "subjid")
```

Identical to `left_join(ae, dm, by = "subjid")` apart from column order, and rare in
practice, because swapping the arguments reads better than making the reader track a
mirror image.
Subject `006` shows up here with `NA` demographics.

### `full_join()`: keep everything from both

```sas
data out;
  merge dm ae;
  by subjid;
run;
```

```{r}
full_join(dm, ae, by = "subjid")
```

Seven rows: every `subjid` from either side. Useful when you are hunting data-quality
problems and do not yet know which side they are on.

### `anti_join()`: left rows with no match on the right

```sas
data out;
  merge dm(in=a) ae(in=b);
  by subjid;
  if a and not b;
run;
```

The QC workhorse. "Who is missing?" is one line in both directions:

```{r}
# Subjects with no adverse events at all
anti_join(dm, ae, by = "subjid")

# AE rows referencing a subject who is not in DM (a data error)
anti_join(ae, dm, by = "subjid")
```

Running both directions after any join is a cheap habit that catches key problems before
they reach an analysis dataset.

### `semi_join()`: filter the left by existence on the right

```sas
proc sql;
  select * from dm where subjid in (select subjid from ae);
quit;
```

```{r}
semi_join(dm, ae, by = "subjid")
```

Three rows, and only `dm`'s columns. Nothing is merged in. Compare with
`inner_join(dm, ae)`, which returned four rows because `001` was duplicated by its two
AEs. **`semi_join()` filters; `inner_join()` filters and merges.** When you only want to
subset, `semi_join()` cannot change your row count behind your back.

## Summary

| R | SAS `DATA` step | SQL |
|---|---|---|
| `inner_join()` | `if a and b;` | `INNER JOIN` |
| `left_join()` | `if a;` | `LEFT JOIN` |
| `right_join()` | `if b;` | `RIGHT JOIN` |
| `full_join()` | no subsetting `if` | `FULL JOIN` |
| `anti_join()` | `if a and not b;` | `WHERE ... NOT IN` |
| `semi_join()` | `if a and b;` then dedupe | `WHERE EXISTS` |

## Keys that are named differently

SAS forces you to rename before merging. R does not:

```{r}
ae_upper <- data.frame(
  SUBJID = c("001", "002", "003"),
  aeterm = c("Headache", "Rash", "Fatigue")
)

left_join(dm, ae_upper, by = c("subjid" = "SUBJID"))
```

Left name first, right name second. The result keeps the left-hand name.

## Joining on more than one key

```{r}
lb <- data.frame(
  subjid = c("001", "001", "002"),
  visit  = c("Baseline", "Week 4", "Baseline"),
  alt    = c(22, 28, 31)
)

visits <- data.frame(
  subjid     = c("001", "001", "002"),
  visit      = c("Baseline", "Week 4", "Baseline"),
  visit_date = as.Date(c("2025-01-10", "2025-02-07", "2025-01-12"))
)

left_join(lb, visits, by = c("subjid", "visit"))
```

A row matches only when **both** keys match, exactly like `BY subjid visit`.

## Row-count hygiene

The most common join bug in clinical data is a right-hand table with duplicate keys,
which quietly multiplies your rows. In SAS you find out by reading the log. In R, state
the shape you expect and let the join fail if the data disagrees:

```{r}
nrow(dm)
nrow(left_join(dm, ae, by = "subjid"))
```

Six rows from five subjects. Here that is correct, because one subject has two AEs, but
you should have to say so:

```{r}
left_join(dm, ae, by = "subjid", relationship = "one-to-many")
```

`relationship =` (dplyr 1.1 and later) takes `"one-to-one"`, `"one-to-many"`,
`"many-to-one"` or `"many-to-many"`, and aborts when the data violates the claim:

```{r}
#| error: true
left_join(dm, ae, by = "subjid", relationship = "one-to-one")
```

That error is the point. Get into the habit of declaring the relationship on every join.
It turns a silent row explosion into a loud failure at the line that caused it.

## Base R: `merge()`

Still common in older code, and the closest single function to a `DATA` step `MERGE`:

```{r}
merge(dm, ae, by = "subjid")                 # inner (the default)
merge(dm, ae, by = "subjid", all.x = TRUE)   # left
merge(dm, ae, by = "subjid", all.y = TRUE)   # right
merge(dm, ae, by = "subjid", all = TRUE)     # full
```

Readable enough, but there is no `merge()` equivalent of `anti_join()` or `semi_join()`,
no `relationship =` check, and `all.x = TRUE` says less at a glance than `left_join`.
Prefer dplyr for new work; recognise `merge()` when you inherit it.
