---
title: "Last observation before exposure"
subtitle: "--LOBXFL flags"
---

::: {.sas-equiv}
**SAS:** last on or before `RFSTDTC`, `--LOBXFL`
:::

`--LOBXFL` is `"Y"` on the last record on or before the reference start, per
entity, and blank otherwise. Build the flag table from qualifying rows, then
join it back. Do not flag in place with a running `last.`.

## Quick reference

{{< include _quickref/lobxfl.md >}}

Worked examples below.

## The example table

Two subjects, two lesions. `RFSTDTC` is on the row.

```{r}
#| label: setup
library(dplyr)

tu <- data.frame(
  USUBJID = c(
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0002", "STD-001-0002"
  ),
  TULNKID = c("T01", "T01", "T01", "T02", "T02", "T02", "T01", "T01"),
  TUDTC   = as.Date(c(
    "2025-05-20",
    "2025-06-03",
    "2025-06-17",
    "2025-05-20",
    "2025-05-20",
    "2025-06-17",
    "2025-06-20",
    "2025-06-27"
  )),
  RFSTDTC = as.Date(c(
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-10", "2025-06-10"
  )),
  TUSTRESC = c("CR", "CR", "SD", "CR", "NE", "PD", "PR", "PR")
)

tu
```

- `STD-001-0001` T01: last on or before `2025-06-03` is `2025-06-03` (on the
  reference date)
- `STD-001-0001` T02: two rows share `2025-05-20`; the later visit is after
  start, so both of those rows are last
- `STD-001-0002` T01: both dates are after `2025-06-10`; nothing qualifies

## Filter, `slice_max()`, join a `"Y"`

Keep rows with `TUDTC <= RFSTDTC`. Per `USUBJID` and `TULNKID`, take the latest
`TUDTC`. Distinct the keys so a tied date does not multiply rows on the join.
Flag `"Y"`, join back, fill the rest with `""`.

```{r}
last_before <- tu |>
  filter(!is.na(TUDTC), !is.na(RFSTDTC), TUDTC <= RFSTDTC) |>
  group_by(USUBJID, TULNKID) |>
  slice_max(TUDTC, n = 1, with_ties = TRUE) |>
  ungroup() |>
  distinct(USUBJID, TULNKID, TUDTC) |>
  mutate(TULOBXFL = "Y")

last_before

tu |>
  left_join(last_before, by = c("USUBJID", "TULNKID", "TUDTC")) |>
  mutate(TULOBXFL = coalesce(TULOBXFL, ""))
```

`STD-001-0001` T01 on `2025-06-03` is `"Y"`. Both `STD-001-0001` T02 rows on
`2025-05-20` are `"Y"` (`with_ties = TRUE`). Post-baseline rows are blank.
`STD-001-0002` is blank on both rows.

::: {.callout-important}
## Blank is `""`, not `NA`

SDTM `--LOBXFL` is `"Y"` or blank. A left join puts `NA` on unmatched rows.
`waldo` and a SAS XPT compare treat `NA` versus `""` as a difference on every
unflagged row. `coalesce(TULOBXFL, "")` is required for that compare.
:::

## Group by the entity

`TULNKID` is the lesion. Last-before-exposure is per lesion, not per subject.
`group_by(USUBJID)` alone selects one date for the subject and can flag the
wrong T02 row, or miss it. Put the IG entity (`TULNKID`, `RSGRPID`, or a record
identifier) in `group_by()`.

## Empty qualifying set stays blank

`STD-001-0002` has no row with `TUDTC <= RFSTDTC`. `last_before` has no keys for
that subject. The left join matches nothing; `coalesce(..., "")` leaves both
rows blank. Do not assign `"Y"` to the earliest post-baseline record.

## `with_ties = TRUE`

`slice_max(..., with_ties = TRUE)` keeps every row that shares the maximum date.
When two records on the same entity fall on the last qualifying day, both get
`"Y"`.

`distinct(USUBJID, TULNKID, TUDTC)` makes the right-hand table unique on the join
keys, so those two left rows each match once. Without `distinct()`, a two-row
tie on the right cartesian-joins against two left rows with the same keys.

`with_ties = FALSE` keeps one row and does not specify which. If the IG requires
exactly one flag per entity on a tied date, pass a tie-breaker to `slice_max()`
(a sequence or a datetime).
