---
title: "Last observation before exposure"
subtitle: "--LOBXFL-style flags"
---

::: {.sas-equiv}
**SAS:** last on or before `RFXSTDTC`, `--LOBXFL`
:::

When a findings domain carries an `--LOBXFL`-style flag, build it in R as:
filter qualifying rows, take the last date per entity, join `"Y"` back, fill
the rest with `""`. Do not flag in place with a running `last.`.

`RFXSTDTC` is first study treatment. `RFSTDTC` is the subject reference start
and may differ. Use `RFXSTDTC` as the cutoff.

## Quick reference

{{< include _quickref/lobxfl.md >}}

Worked examples below.

## The example table

Two subjects, two vital-sign tests. `RFXSTDTC` is on the row.

```{r}
#| label: setup
library(dplyr)

vs <- data.frame(
  USUBJID = c(
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0002", "STD-001-0002"
  ),
  VSTESTCD = c("SYSBP", "SYSBP", "SYSBP", "DIABP", "DIABP", "DIABP", "SYSBP", "SYSBP"),
  VSDTC   = as.Date(c(
    "2025-05-20",
    "2025-06-03",
    "2025-06-17",
    "2025-05-20",
    "2025-05-20",
    "2025-06-02",
    "2025-06-20",
    "2025-06-27"
  )),
  RFXSTDTC = as.Date(c(
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-10", "2025-06-10"
  )),
  VSSTRESN = c(122, 118, 130, 80, 78, NA_real_, 125, 122)
)

vs
```

- `STD-001-0001` SYSBP: last on or before `2025-06-03` with a result is
  `2025-06-03` (same calendar day as first dose)
- `STD-001-0001` DIABP: `2025-06-02` has no result, so it does not qualify;
  two rows share `2025-05-20` and both get the flag
- `STD-001-0002` SYSBP: both dates are after `2025-06-10`; nothing qualifies

## Filter, `slice_max()`, join a `"Y"`

Drop missing dates and missing or not-done results, then keep
`VSDTC <= RFXSTDTC`. Per `USUBJID` and `VSTESTCD`, take the latest `VSDTC`.
Distinct the keys so a tied date does not multiply rows on the join. Flag
`"Y"`, join back, fill the rest with `""`.

```{r}
last_before <- vs |>
  filter(!is.na(VSDTC), !is.na(RFXSTDTC), !is.na(VSSTRESN), VSDTC <= RFXSTDTC) |>
  group_by(USUBJID, VSTESTCD) |>
  slice_max(VSDTC, n = 1, with_ties = TRUE) |>
  ungroup() |>
  distinct(USUBJID, VSTESTCD, VSDTC) |>
  mutate(VSLOBXFL = "Y")

last_before

vs |>
  left_join(last_before, by = c("USUBJID", "VSTESTCD", "VSDTC")) |>
  mutate(VSLOBXFL = coalesce(VSLOBXFL, ""))
```

`STD-001-0001` SYSBP on `2025-06-03` is `"Y"`. Both `STD-001-0001` DIABP rows on
`2025-05-20` are `"Y"` (`with_ties = TRUE`). The DIABP row with a missing result
stays blank. Post-baseline rows are blank. `STD-001-0002` is blank on both rows.

These dates have no times. A visit on the same calendar day as first dose
cannot be ordered by datetime. The study rule must say whether that day counts
as before, after, or is decided by visit or timepoint. The pattern above treats
same calendar day as on or before (`<=`).

::: {.callout-important}
## Blank is `""`, not `NA`

`--LOBXFL` is `"Y"` or blank. A left join puts `NA` on unmatched rows.
`waldo` and a SAS XPT compare treat `NA` versus `""` as a difference on every
unflagged row. `coalesce(VSLOBXFL, "")` is required for that compare.
:::

## Group by the entity

Last-before-exposure is per test here (`VSTESTCD`), not per subject.
`group_by(USUBJID)` alone selects one date for the subject and can flag the
wrong DIABP row, or miss it. Put the finding key in `group_by()`.

## Empty qualifying set stays blank

`STD-001-0002` has no row with `VSDTC <= RFXSTDTC`. `last_before` has no keys for
that subject. The left join matches nothing; `coalesce(..., "")` leaves both
rows blank. Do not assign `"Y"` to the earliest post-baseline record.

## `with_ties = TRUE`

`slice_max(..., with_ties = TRUE)` keeps every row that shares the maximum date.
When two records on the same entity fall on the last qualifying day, both get
`"Y"`.

`distinct(USUBJID, VSTESTCD, VSDTC)` makes the right-hand table unique on the
join keys, so those two left rows each match once. Without `distinct()`, a
two-row tie on the right cartesian-joins against two left rows with the same
keys.

`with_ties = FALSE` keeps one row and does not specify which. If the study
requires exactly one flag per entity on a tied date, pass a tie-breaker to
`slice_max()` (a sequence or a datetime).
