---
title: "Last observation before exposure"
subtitle: "--LOBXFL flags"
---

::: {.sas-equiv}
**SAS:** last on or before `RFSTDTC`, `--LOBXFL`
:::

`--LOBXFL` is `"Y"` on the last record on or before the reference start, per entity,
and blank everywhere else. Build the flag table from the qualifying rows, then join
it back. Do not try to flag in place with a running `last.`.

## Quick reference

{{< include _quickref/lobxfl.md >}}

The rest of this page is the same material worked through slowly, with runnable examples
and the output they produce. Skip it if the block above was what you came for.

## The example table

Two subjects, two lesions. `RFSTDTC` is already on the row.

```{r}
#| label: setup
library(dplyr)

tu <- data.frame(
  USUBJID = c(
    "XYZ-101-0001", "XYZ-101-0001", "XYZ-101-0001",
    "XYZ-101-0001", "XYZ-101-0001", "XYZ-101-0001",
    "XYZ-101-0002", "XYZ-101-0002"
  ),
  TULNKID = c("T01", "T01", "T01", "T02", "T02", "T02", "T01", "T01"),
  TUDTC   = as.Date(c(
    "2025-05-20",
    "2025-06-03",
    "2025-06-17",
    "2025-05-20",
    "2025-05-20",
    "2025-06-17",
    "2025-06-20",
    "2025-06-27"
  )),
  RFSTDTC = as.Date(c(
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-10", "2025-06-10"
  )),
  TUSTRESC = c("CR", "CR", "SD", "CR", "NE", "PD", "PR", "PR")
)

tu
```

- `0001` T01: last on or before `2025-06-03` is `2025-06-03` (on the reference date)
- `0001` T02: two rows share `2025-05-20`, and the later visit is after start, so both
  of those rows qualify as last
- `0002` T01: both dates are after `2025-06-10`, so nothing qualifies

## Filter, `slice_max()`, join a `"Y"`

Keep rows on or before `RFSTDTC`. Per `USUBJID` and `TULNKID`, take the latest
`TUDTC`. Distinct the keys so a tied date does not explode the join. Flag `"Y"`,
join back, and fill the rest with `""`.

```{r}
last_before <- tu |>
  filter(!is.na(TUDTC), !is.na(RFSTDTC), TUDTC <= RFSTDTC) |>
  group_by(USUBJID, TULNKID) |>
  slice_max(TUDTC, n = 1, with_ties = TRUE) |>
  ungroup() |>
  distinct(USUBJID, TULNKID, TUDTC) |>
  mutate(TULOBXFL = "Y")

last_before

tu |>
  left_join(last_before, by = c("USUBJID", "TULNKID", "TUDTC")) |>
  mutate(TULOBXFL = coalesce(TULOBXFL, ""))
```

`0001` T01 on 3 June is `"Y"`. Both `0001` T02 rows on 20 May are `"Y"` because
`with_ties = TRUE`. Post-baseline rows are blank. `0002` is blank on both rows.

::: {.callout-important}
## Blank is `""`, not `NA`

SDTM `--LOBXFL` is `"Y"` or blank. A left join puts `NA` on rows with no match.
`waldo` and a SAS XPT compare will report `NA` versus `""` as a real difference on
every unflagged row. `coalesce(TULOBXFL, "")` is part of the flag, not polish.
:::

## Group by the entity

`TULNKID` is the lesion. Last-before-exposure is per lesion, not per subject.
`group_by(USUBJID)` alone would pick one date for the whole subject and flag the
wrong T02 row, or miss it. Whatever the IG calls the entity (`TULNKID`, `RSGRPID`,
a record identifier) goes in the `group_by()`.

## Empty qualifying set stays blank

`0002` has no row with `TUDTC <= RFSTDTC`. `last_before` has no `0002` keys. The
left join therefore matches nothing, and `coalesce(..., "")` leaves both rows
blank. Do not invent a `"Y"` on the earliest post-baseline record.

## `with_ties = TRUE`

`slice_max(..., with_ties = TRUE)` keeps every row that shares the maximum date.
That is what you want when two records on the same entity fall on the same last
qualifying day: both get `"Y"`.

`distinct(USUBJID, TULNKID, TUDTC)` then makes the right-hand table unique on the
join keys, so those two left rows each match once. Skip `distinct()` and a two-row
tie on the right cartesian-joins against two left rows with the same keys.

`with_ties = FALSE` drops down to one row and does not say which. If the IG wants
exactly one flag per entity on a tied date, add a tie-breaker to `slice_max()`
(a sequence, a datetime) rather than leaving the choice implicit.
