---
title: "Mutate"
subtitle: "Create or change columns"
---

::: {.sas-equiv}
**SAS:** DATA step assignment (`new = ...;`, overwrite existing)
:::

`dplyr::mutate()` adds or replaces columns. One call can set several columns.
The result is a new data frame; the input is unchanged unless you overwrite it.

## Quick reference

{{< include _quickref/mutate.md >}}

Worked examples below.

## Add a column

```{r}
#| label: add
library(dplyr)

dm <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002"),
  AGE = c(54L, 61L)
)

dm |>
  mutate(AGEU = "YEARS")
```

SAS:

```sas
ageu = "YEARS";
```

## Overwrite a column

```{r}
#| label: overwrite
dm |>
  mutate(AGE = AGE + 1L)
```

Same name on the left replaces the column. Order matters when later columns
use earlier ones in the same `mutate()`.

## Several columns

```{r}
#| label: several
dm |>
  mutate(
    AGEU = "YEARS",
    AGEGR1 = if_else(AGE < 65L, "<65", ">=65")
  )
```

One `mutate()` is one DATA step block of assignments. Prefer that over chained
one-column `mutate()` calls when the columns belong together.

::: {.callout-important}
## Not in place

`mutate(dm, ...)` alone does nothing lasting. Use `dm <- dm |> mutate(...)` or
keep piping into the next verb.
:::
