---
title: "Partial dates"
subtitle: "Impute year-only or year-month to a full date"
---

::: {.sas-equiv}
**SAS:** partial `--DTC` / character dates; impute with `INTNX` beginning/end or a fixed mid day
:::

Clinical data often arrive as incomplete ISO-like strings: year only (`2025`)
or year-month (`2024-04`). Derivation rules usually push these to the first,
middle, or last day of the known period. Month bounds matter most for
year-month; year-only is the same idea at year scale.

Always follow the study imputation convention and record how far you imputed
(for example CDISC `--DTF`) when required. Examples below use generic
`STD-001-*` subjects and one explicit mid rule.

## Quick reference

{{< include _quickref/partial-dates.md >}}

Worked examples below.

## Year-month to first, mid, last

Three partial year-month values, including February in a leap year.

```{r}
#| label: ym
library(dplyr)
library(lubridate)

partial_ym <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0001", "STD-001-0002"),
  DTC_YM  = c("2024-02", "2024-03", "2025-12")
)

partial_ym |>
  mutate(
    START = ymd(paste0(DTC_YM, "-01")),
    MID   = ymd(paste0(DTC_YM, "-15")),
    END   = (ceiling_date(ymd(paste0(DTC_YM, "-01")), "month") - days(1)) |>
      as.Date()
  )
```

| `DTC_YM` | first | mid (day 15) | last |
|---|---|---|---|
| 2024-02 | 2024-02-01 | 2024-02-15 | 2024-02-29 |
| 2024-03 | 2024-03-01 | 2024-03-15 | 2024-03-31 |
| 2025-12 | 2025-12-01 | 2025-12-15 | 2025-12-31 |

First day matches SAS `INTNX('month', ..., 0, 'beginning')` on a real date in
that month. Last day matches `INTNX(..., 'end')`. Day 15 is a common mid-month
convention; some protocols use other mid rules. Pick one and keep it consistent.

::: {.callout-important}
## Month end is not bare `ceiling_date`

`ceiling_date(start, "month")` is the first day of the *next* month. Last day of
the current month is that value minus one day.
:::

## Year-only to first, mid, last

Year-only strings are easier: fixed calendar anchors.

```{r}
#| label: year
partial_y <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002"),
  DTC_Y   = c("2024", "2025")
)

partial_y |>
  mutate(
    START = ymd(paste0(DTC_Y, "-01-01")),
    MID   = ymd(paste0(DTC_Y, "-06-30")),
    END   = ymd(paste0(DTC_Y, "-12-31"))
  )
```

| `DTC_Y` | first | mid (30 Jun) | last |
|---|---|---|---|
| 2024 | 2024-01-01 | 2024-06-30 | 2024-12-31 |
| 2025 | 2025-01-01 | 2025-06-30 | 2025-12-31 |

30 June is a common mid-year convention (half-year boundary). Some conventions
use 1 July or day 183. Document the choice the same way you document day 15 for
months.

SAS analogs: `INTNX('year', ..., 0, 'beginning')` / `'end'`, or a fixed mid date
constructed from the year.

## Choosing first vs last vs mid

| Situation | Typical choice |
|---|---|
| Conservative "earliest possible" onset | first of period |
| Conservative "latest possible" | last of period |
| Neutral / midpoint rule in SAP | mid (day 15 / 30 Jun or protocol rule) |

Do not invent a mid rule in production code. Read the SAP or derivation
specification. When CDISC date imputation flags apply, set `--DTF` to show which
parts were imputed (for example `D` when day was imputed, `M` when month and day
were imputed). Exact flag codes follow the standard and the study metadata.

Missing or unusable partial strings stay missing after failed parses. Do not
silently coerce blank `--DTC` to period start.
