---
title: "Select and rename"
subtitle: "Keep, drop, and rename columns"
---

::: {.sas-equiv}
**SAS:** `KEEP`, `DROP`, `RENAME`
:::

`select()` chooses columns. `rename()` changes names without dropping others.

## Quick reference

{{< include _quickref/select-rename.md >}}

Worked examples below.

## Keep listed columns

```{r}
#| label: keep
library(dplyr)

ae <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0001"),
  AETERM = c("NAUSEA", "HEADACHE"),
  AESEV = c("MILD", "MODERATE"),
  TEMP = c(1L, 2L)
)

ae |>
  select(USUBJID, AETERM, AESEV)
```

SAS: `keep usubjid aeterm aesev;`

## Drop columns

```{r}
#| label: drop
ae |>
  select(-TEMP)
```

SAS: `drop temp;`

## Rename

```{r}
#| label: rename
ae |>
  rename(AEDECOD = AETERM)
```

::: {.callout-important}
## Name order differs from SAS

SAS: `rename aeterm = aedecod;` (old = new).  
R: `rename(AEDECOD = AETERM)` (new = old).
:::

`select(AEDECOD = AETERM, everything())` can rename and keep the rest in one
step. Plain `rename()` is clearer when you only change names.
