---
title: "Sequence numbers"
subtitle: "--SEQ within a subject"
---

::: {.sas-equiv}
**SAS:** `PROC SORT` then `_N_` within `BY`, `--SEQ`
:::

`--SEQ` is a within-subject row number after a defined order. In R that is
`arrange()`, then `group_by(USUBJID)`, then `row_number()`. Skip the sort and you
still get a sequence. It is just not the sequence the IG asked for.

## Quick reference

{{< include _quickref/seq.md >}}

The rest of this page is the same material worked through slowly, with runnable examples
and the output they produce. Skip it if the block above was what you came for.

## The example table

Tumor results in arrival order, which is not the order `TUSEQ` should use.

```{r}
#| label: setup
library(dplyr)

tu <- data.frame(
  USUBJID  = c(
    "XYZ-101-0001", "XYZ-101-0001", "XYZ-101-0001",
    "XYZ-101-0001", "XYZ-101-0002"
  ),
  TUEVAL   = c(
    "INVESTIGATOR", "INVESTIGATOR", "INVESTIGATOR",
    "INDEPENDENT ASSESSOR", "INVESTIGATOR"
  ),
  TULNKID  = c("T01", "T01", "T02", "T01", "T01"),
  VISITNUM = c(3, 1, 1, 1, 2),
  TUDY     = c(15, 1, 1, 1, 8),
  TUSTRESC = c("SD", "CR", "CR", "CR", "PR")
)

tu
```

`0001` has investigator lesions T01 and T02, plus one independent-assessor row.
Visit 3 arrived first. The IG order is evaluator, lesion, visit, study day.

## Wrong: `row_number()` without `arrange()`

```{r}
tu |>
  group_by(USUBJID) |>
  mutate(TUSEQ = row_number()) |>
  ungroup()
```

`0001` is numbered 1..4 in file order. Visit 3 is `TUSEQ` 1 because it was on top.
That is `_N_` without the `PROC SORT`.

## Right: `arrange()`, then number

```sas
proc sort data=tu;
  by usubjid tueval tulnkid visitnum tudy;
run;

data tu;
  set tu;
  by usubjid;
  if first.usubjid then tuseq = 0;
  tuseq + 1;
run;
```

```{r}
tu |>
  arrange(USUBJID, TUEVAL, TULNKID, VISITNUM, TUDY) |>
  group_by(USUBJID) |>
  mutate(TUSEQ = as.double(row_number())) |>
  ungroup()
```

`INDEPENDENT ASSESSOR` sorts before `INVESTIGATOR`, so it is `TUSEQ` 1. Then
investigator T01 visit 1, T01 visit 3, T02 visit 1. `TULNKID` comes before
`VISITNUM`, so T01 visit 3 is still ahead of T02 visit 1. `0002` restarts at 1.

::: {.callout-important}
## `row_number()` numbers the current order

It does not infer a clinical order. If you `group_by(USUBJID)` and number, you get
the order the frame happens to be in. Sort explicitly. Put every key the IG names
into `arrange()`, including the tie-breakers.
:::

## Tie-breakers

Two rows can share a study day. `arrange(USUBJID, TUDY)` alone then leaves their
relative order to whatever came before. Add the keys that make a row unique, in
the IG's order: `TUEVAL`, `TULNKID`, `VISITNUM`, `TUDY`. If two rows still tie,
add a source sequence or a datetime. Do not leave a tie unnamed.

## `as.double()` for XPT

SAS numeric is double. `row_number()` returns integer. `haven` / `xportr` will
write that as an integer column unless you cast it. SDTM `--SEQ` in a SAS XPT is
double, so:

```{r}
tu |>
  arrange(USUBJID, TUEVAL, TULNKID, VISITNUM, TUDY) |>
  group_by(USUBJID) |>
  mutate(TUSEQ = as.double(row_number())) |>
  ungroup() |>
  pull(TUSEQ) |>
  typeof()
```

`double`. Match the SAS type now, not after the first failed compare against production.
