---
title: "Sequence numbers"
subtitle: "--SEQ within a subject"
---

::: {.sas-equiv}
**SAS:** `PROC SORT` then `first.` and a retained counter, `--SEQ`
:::

`--SEQ` is a within-subject row number after a defined order. In R: `arrange()`,
then `group_by(USUBJID)`, then `row_number()`. Without `arrange()`, the numbers
follow the current row order, not the study keys.

## Quick reference

{{< include _quickref/seq.md >}}

Worked examples below.

## The example table

Tumor results in input order, which is not the `TUSEQ` order.

```{r}
#| label: setup
library(dplyr)

tu <- data.frame(
  USUBJID  = c(
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0001", "STD-001-0002"
  ),
  TUEVAL   = c(
    "INVESTIGATOR", "INVESTIGATOR", "INVESTIGATOR",
    "INDEPENDENT ASSESSOR", "INVESTIGATOR"
  ),
  TULNKID  = c("T01", "T01", "T02", "T01", "T01"),
  VISITNUM = c(3, 1, 1, 1, 2),
  TUDY     = c(15, 1, 1, 1, 8),
  TUSTRESC = c("SD", "CR", "CR", "CR", "PR")
)

tu
```

`STD-001-0001` has investigator lesions T01 and T02, and one independent-assessor
row. `VISITNUM` 3 is the first input row. Sponsor order used here: evaluator,
lesion, visit, study day.

## Wrong: `row_number()` without `arrange()`

```{r}
tu |>
  group_by(USUBJID) |>
  mutate(TUSEQ = row_number()) |>
  ungroup()
```

`STD-001-0001` is numbered 1 to 4 in file order. `VISITNUM` 3 is `TUSEQ` 1.
The SAS counter restarts per subject too, but only after you sort.

## Right: `arrange()`, then number

```sas
proc sort data=tu;
  by usubjid tueval tulnkid visitnum tudy;
run;

data tu;
  set tu;
  by usubjid;
  if first.usubjid then tuseq = 0;
  tuseq + 1;
run;
```

```{r}
tu |>
  arrange(USUBJID, TUEVAL, TULNKID, VISITNUM, TUDY) |>
  group_by(USUBJID) |>
  mutate(TUSEQ = as.double(row_number())) |>
  ungroup()
```

`INDEPENDENT ASSESSOR` sorts before `INVESTIGATOR` and is `TUSEQ` 1. Then
investigator T01 visit 1, T01 visit 3, T02 visit 1. `TULNKID` precedes
`VISITNUM`, so T01 visit 3 precedes T02 visit 1. `STD-001-0002` restarts at 1.

`_N_` does not restart within `BY`. The SAS analogue of `group_by(USUBJID)` then
`row_number()` is the `first.` plus retained counter above, not `_N_`.

::: {.callout-important}
## `row_number()` numbers the current order

It does not infer a clinical order. `group_by(USUBJID)` then `row_number()` uses
the frame's current order. Sort first. Include every study key in `arrange()`,
including tie-breakers.
:::

## Tie-breakers

Two rows can share a study day. `arrange(USUBJID, TUDY)` alone leaves their
relative order undefined. Add keys that make a row unique, in the sponsor-defined
order: `TUEVAL`, `TULNKID`, `VISITNUM`, `TUDY`. If a tie remains, add a source
sequence or a datetime.

## `as.double()` for R-side compares

XPT has numeric and character only. Integer is not a distinct XPT type.
`row_number()` returns integer in R. SAS numeric is double. `waldo` and
`diffdf` report that as a type difference even when the values match. Cast when
you need the R objects to compare equal:

```{r}
tu |>
  arrange(USUBJID, TUEVAL, TULNKID, VISITNUM, TUDY) |>
  group_by(USUBJID) |>
  mutate(TUSEQ = as.double(row_number())) |>
  ungroup() |>
  pull(TUSEQ) |>
  typeof()
```

`double`. The cast is for R-side compares against SAS numeric, not because XPT
stores a separate integer type.
