---
title: "Study day"
subtitle: "Date subtraction for --DY"
---

::: {.sas-equiv}
**SAS:** date minus date, `INTCK('day', ...)`, CDISC `--DY`
:::

CDISC `--DY` has no day 0. The reference start date is day 1. Dates before it stay
negative and do not shift. Raw subtraction is the wrong answer on and after the
reference date.

## Quick reference

{{< include _quickref/study-day.md >}}

The rest of this page is the same material worked through slowly, with runnable examples
and the output they produce. Skip it if the block above was what you came for.

## Raw diff versus `--DY`

Five lab dates around one reference start. First two are before it, then the
reference day itself, then two after.

```{r}
#| label: setup
library(dplyr)

lb <- data.frame(
  USUBJID = c(
    "XYZ-101-0001", "XYZ-101-0001", "XYZ-101-0001",
    "XYZ-101-0001", "XYZ-101-0001"
  ),
  LBDT   = as.Date(c(
    "2025-06-01", "2025-06-02", "2025-06-03",
    "2025-06-04", "2025-06-05"
  )),
  RFSTDT = as.Date("2025-06-03")
)

lb |>
  mutate(
    RAW  = as.integer(LBDT - RFSTDT),
    LBDY = if_else(
      LBDT < RFSTDT,
      as.integer(LBDT - RFSTDT),
      as.integer(LBDT - RFSTDT) + 1L
    )
  )
```

| `LBDT` vs `RFSTDT` | raw | `LBDY` |
|---|---|---|
| two days before | -2 | -2 |
| one day before | -1 | -1 |
| on the reference date | 0 | 1 |
| one day after | 1 | 2 |
| two days after | 2 | 3 |

The day before and the day after the reference are not symmetric. That is the rule.
SAS date arithmetic gives you the raw column. `--DY` is the `LBDY` column.

::: {.callout-important}
## There is no day 0

`date - rfstdt` on the reference date is 0. `--DY` on that date is 1. Every date on
or after the reference is shifted by +1. Dates before it are left as a negative
offset. If you ship the raw diff, day 1 of treatment is labelled 0 and every
subsequent day is off by one.
:::

## Compact form

The branch is a +1 when `date >= rfstdt`, and +0 otherwise:

```{r}
lb |>
  mutate(
    LBDY = as.integer(LBDT - RFSTDT) + as.integer(LBDT >= RFSTDT)
  )
```

`as.integer(TRUE)` is 1. `as.integer(FALSE)` is 0. Same values as the `if_else()`
form. `date - rfstdt` is a `difftime`; `as.integer()` turns it into a number before
you add 1.

## A subject table

Two subjects. `0002` starts on `2025-06-10` and has one lab on that day. Missing
`LBDT` stays missing.

```{r}
lb2 <- data.frame(
  USUBJID = c(
    "XYZ-101-0001", "XYZ-101-0001", "XYZ-101-0001",
    "XYZ-101-0002", "XYZ-101-0002"
  ),
  LBDT = as.Date(c(
    "2025-06-02", "2025-06-03", "2025-06-04",
    "2025-06-10", NA
  )),
  RFSTDT = as.Date(c(
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-10", "2025-06-10"
  ))
)

lb2 |>
  mutate(
    LBDY = as.integer(LBDT - RFSTDT) + as.integer(LBDT >= RFSTDT)
  )
```

`0001` has study days -1, 1, 2. `0002` has study day 1, then `NA`. Do not fill
missing dates with 0. CDISC leaves `--DY` missing when either date is missing.

`INTCK('day', rfstdt, date)` in SAS is the raw interval, not `--DY`. Add the same
+1 on and after the reference date there too.
