---
title: "Study day"
subtitle: "Date subtraction for --DY"
---

::: {.sas-equiv}
**SAS:** date minus date, `INTCK('day', ...)`, CDISC `--DY`
:::

CDISC `--DY` has no day 0. The reference start date is day 1. Dates before the
reference stay negative and are not shifted. Raw subtraction is not `--DY` on or
after the reference date.

## Quick reference

{{< include _quickref/study-day.md >}}

Worked examples below.

## Raw diff versus `--DY`

Five lab dates around one reference start: two before, the reference date, two
after.

```{r}
#| label: setup
library(dplyr)

lb <- data.frame(
  USUBJID = c(
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0001", "STD-001-0001"
  ),
  LBDT   = as.Date(c(
    "2025-06-01", "2025-06-02", "2025-06-03",
    "2025-06-04", "2025-06-05"
  )),
  RFSTDT = as.Date("2025-06-03")
)

lb |>
  mutate(
    RAW  = as.integer(LBDT - RFSTDT),
    LBDY = if_else(
      LBDT < RFSTDT,
      as.integer(LBDT - RFSTDT),
      as.integer(LBDT - RFSTDT) + 1L
    )
  )
```

| `LBDT` vs `RFSTDT` | raw | `LBDY` |
|---|---|---|
| two days before | -2 | -2 |
| one day before | -1 | -1 |
| on the reference date | 0 | 1 |
| one day after | 1 | 2 |
| two days after | 2 | 3 |

The day before and the day after the reference are not symmetric. SAS date
arithmetic returns the raw column. `--DY` is the `LBDY` column.

::: {.callout-important}
## There is no day 0

`date - rfstdt` on the reference date is 0. `--DY` on that date is 1. Every date
on or after the reference is shifted by +1. Dates before the reference remain a
negative offset. The raw diff labels the first treatment day as 0 and offsets
every later day by one.
:::

## Compact form

Add 1 when `date >= rfstdt`, otherwise add 0:

```{r}
lb |>
  mutate(
    LBDY = as.integer(LBDT - RFSTDT) + as.integer(LBDT >= RFSTDT)
  )
```

`as.integer(TRUE)` is 1. `as.integer(FALSE)` is 0. Same values as `if_else()`.
`date - rfstdt` is a `difftime`; `as.integer()` converts it before the +1.

## A subject table

Two subjects. `STD-001-0002` starts on `2025-06-10` and has one lab on that day.
Missing `LBDT` remains missing.

```{r}
lb2 <- data.frame(
  USUBJID = c(
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0002", "STD-001-0002"
  ),
  LBDT = as.Date(c(
    "2025-06-02", "2025-06-03", "2025-06-04",
    "2025-06-10", NA
  )),
  RFSTDT = as.Date(c(
    "2025-06-03", "2025-06-03", "2025-06-03",
    "2025-06-10", "2025-06-10"
  ))
)

lb2 |>
  mutate(
    LBDY = as.integer(LBDT - RFSTDT) + as.integer(LBDT >= RFSTDT)
  )
```

`STD-001-0001` study days: -1, 1, 2. `STD-001-0002`: 1, then `NA`. Do not replace
missing dates with 0. CDISC leaves `--DY` missing when either date is missing.

`INTCK('day', rfstdt, date)` in SAS is the raw interval, not `--DY`. Apply the
same +1 on and after the reference date.
