---
title: "Substrings"
subtitle: "Extract part of a character value"
---

::: {.sas-equiv}
**SAS:** `SUBSTR(string, start, length)`, `SUBSTRN`
:::

SAS `SUBSTR` takes a length. R `substr()` takes a stop position. The three-argument
shape is the same; the last argument is not. A SAS length passed as `stop` can
return `""` with no error.

## Quick reference

{{< include _quickref/substr.md >}}

Worked examples below.

## The example values

`USUBJID` values of the form study-site-subject, and visit codes of the form
`CYCnDn`.

```{r}
#| label: setup
library(dplyr)

dm <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002", "STD-002-0001"),
  VISIT   = c("CYC1D1", "CYC2D1", "CYC1D8")
)

dm
```

`STD-001-0001` is 12 characters. Positions 5-7 are the site, `001`.
`CYC1D1`: cycle at position 4, day from position 6.

## The empty-range mistake

SAS `SUBSTR(usubjid, 5, 3)` means start at 5, length 3: the site.

```sas
site = substr(usubjid, 5, 3);
```

R `substr(usubjid, 5, 3)` means start at 5, stop at 3. Stop is before start, so
the range is empty:

```{r}
substr(dm$USUBJID, 5, 3)
```

Three empty strings. No warning.

Use the end position, or write the SAS length as `start + len - 1`:

```{r}
substr(dm$USUBJID, 5, 7)
substr(dm$USUBJID, 5, 5 + 3 - 1)
```

::: {.callout-important}
## The third argument is a stop position, not a length

`substr(x, start, stop)` keeps characters from `start` through `stop` inclusive.
A SAS length with `stop < start` returns `""` on every row. Count the end
position, or write `start + len - 1`.
:::

## Visit codes: `CYC1D1`

Cycle is one character at position 4. Day starts at position 6 and continues to
the end of the string, so `CYC1D8` and `CYC1D15` both work with `nchar()`.

```{r}
dm |>
  mutate(
    CYCLE = substr(VISIT, 4, 4),
    DAY   = substr(VISIT, 6, nchar(VISIT))
  )
```

SAS for the cycle is `SUBSTR(visit, 4, 1)`. In R that is `substr(visit, 4, 4)`,
not `substr(visit, 4, 1)`.

## `substring()` and `str_sub()`

Base R `substring()` uses the same start-and-stop model as `substr()`. Arguments
recycle when lengths differ. Assignment is supported
(`substring(x, 5, 7) <- "000"`). For extraction, `substr()` is sufficient.

`stringr::str_sub()` also takes end positions, not a length. `str_sub(x, 5, 7)`
matches `substr(x, 5, 7)`. Negative indices count from the end. It is not SAS
`SUBSTR(x, 5, 3)`. For a length API, compute the stop position.
