---
title: "Substrings"
subtitle: "Extract part of a character value"
---

::: {.sas-equiv}
**SAS:** `SUBSTR(string, start, length)`, `SUBSTRN`
:::

SAS `SUBSTR` takes a length. R `substr()` takes a stop position. Same three-argument
shape, different meaning for the last one. Mixing them up returns `""` and does not
error.

## Quick reference

{{< include _quickref/substr.md >}}

The rest of this page is the same material worked through slowly, with runnable examples
and the output they produce. Skip it if the block above was what you came for.

## The example values

Study-site-subject identifiers, and a few visit codes of the `CYCnDn` form.

```{r}
#| label: setup
library(dplyr)

dm <- data.frame(
  USUBJID = c("XYZ-101-0001", "XYZ-101-0002", "XYZ-102-0001"),
  VISIT   = c("CYC1D1", "CYC2D1", "CYC1D8")
)

dm
```

`XYZ-101-0001` is 12 characters. Positions 5-7 are the site, `101`.
`CYC1D1` is cycle 1 day 1: cycle at position 4, day from position 6.

## The empty-range mistake

SAS `SUBSTR(usubjid, 5, 3)` means start at 5, take 3 characters: the site.

```sas
site = substr(usubjid, 5, 3);
```

R `substr(usubjid, 5, 3)` means start at 5, stop at 3. Stop is before start, so the
range is empty:

```{r}
substr(dm$USUBJID, 5, 3)
```

Three empty strings. No warning. The call looks like SAS and does nothing.

Fix it by giving the end position, or by writing the SAS length as `start + len - 1`:

```{r}
substr(dm$USUBJID, 5, 7)
substr(dm$USUBJID, 5, 5 + 3 - 1)
```

::: {.callout-important}
## The third argument is a stop position, not a length

`substr(x, start, stop)` keeps characters from `start` through `stop` inclusive.
If you pass a SAS length and `stop < start`, R returns `""` for every row. Count
the end position, or write `start + len - 1`. The call will not tell you that you
translated the signature wrong.
:::

## Visit codes: `CYC1D1`

Cycle is one character at position 4. Day starts at position 6 and runs to the end,
so `CYC1D8` and `CYC1D15` both work if you stop at `nchar()`.

```{r}
dm |>
  mutate(
    CYCLE = substr(VISIT, 4, 4),
    DAY   = substr(VISIT, 6, nchar(VISIT))
  )
```

SAS for the cycle would be `SUBSTR(visit, 4, 1)`. In R that is `substr(visit, 4, 4)`,
not `substr(visit, 4, 1)`.

## `substring()` and `str_sub()`

Base R also has `substring()`. Same start-and-stop model as `substr()`. Arguments
recycle when lengths differ, and you can assign into it (`substring(x, 5, 7) <- "000"`).
For ordinary extraction, `substr()` is enough.

`stringr::str_sub()` also takes end positions, not a length. `str_sub(x, 5, 7)` matches
`substr(x, 5, 7)`. Negative indices count from the end. It does not restore SAS
`SUBSTR(x, 5, 3)`. If you want a length API, compute the stop yourself.
