---
title: "Change text case"
subtitle: "toupper() / tolower() / toTitleCase()"
---

::: {.sas-equiv}
**SAS:** `UPCASE`, `LOWCASE`, `PROPCASE`
:::

`toupper()` and `tolower()` match `UPCASE` and `LOWCASE` on ASCII letters `A-Z`
and `a-z`. Digits and symbols are unchanged in both languages. Base R has no
`PROPCASE`. The house stand-in for plain words is
`tools::toTitleCase(tolower(x))`. It is not the same function.

## Quick reference

{{< include _quickref/text-case.md >}}

Worked examples below.

## `USUBJID` and `ARM`

```{r}
#| label: setup
library(dplyr)

ae <- data.frame(
  USUBJID = c("std-001-0001", "STD-001-0002"),
  ARM = c("Placebo", "Active"),
  AETERM = c("HEADACHE OF MILD SEVERITY", "Nausea")
)

ae |>
  mutate(
    USUBJID = toupper(USUBJID),
    ARM_UP = toupper(ARM),
    ARM_LO = tolower(ARM)
  )
```

SAS:

```sas
usubjid = upcase(usubjid);
arm_up = upcase(arm);
arm_lo = lowcase(arm);
```

`toupper()` changes `std-001-0001` to `STD-001-0001`. Hyphens and digits stay.
`ARM_UP` is `PLACEBO` and `ACTIVE`. `ARM_LO` is `placebo` and `active`.

## Proper case

`PROPCASE` lowercases the argument, then capitalizes the first character of
each word. The default word breaks are blank, `/`, `-`, `(`, `.`, and tab.
SAS documents `PROPCASE("INTRODUCTION TO THE SCIENCE OF ASTRONOMY")` as
`Introduction To The Science Of Astronomy`. Words such as `To` and `Of` are
capitalized.

```{r}
#| label: title
ae |>
  mutate(AETERM_TC = tools::toTitleCase(tolower(AETERM)))
```

`tools::toTitleCase(tolower(AETERM))` returns `Headache of Mild Severity` and
`Nausea`. `of` stays lowercase. `PROPCASE("HEADACHE OF MILD SEVERITY")` returns
`Headache Of Mild Severity`.

`tolower()` first is required. `toTitleCase()` leaves an all-caps word unchanged:

```{r}
#| label: caps
tools::toTitleCase("HEADACHE")
tools::toTitleCase(tolower("HEADACHE"))
```

`HEADACHE`, then `Headache`. SAS `propcase("HEADACHE")` is `Headache`.

`.` is a `PROPCASE` word break and is not a `toTitleCase()` word break.
SAS documents `PROPCASE("WINSTON-SALEM, N.C.")` as `Winston-Salem, N.C.`.

```{r}
#| label: period
tools::toTitleCase(tolower("WINSTON-SALEM, N.C."))
```

`N.C.` becomes `N.c.`.

::: {.callout-important}
## `toTitleCase()` is not `PROPCASE`

`PROPCASE` capitalizes every word and splits on `.`.
`tools::toTitleCase()` leaves English function words (`of`, `the`, `to`, and
others) lowercase unless they lead the string, and it does not split on `.`.
Call `tolower()` first or an all-caps word is returned unchanged.
:::

## Factors

```{r}
#| label: factor
arm <- factor(c("Placebo", "Active"))
toupper(arm)
class(toupper(arm))
```

The result is character, the uppercased labels, not a factor. SAS character
values are not factors. Wrap `factor()` again if a later step requires one.

`toupper(NA)` is `NA`. `toupper("")` is `""`. SAS `UPCASE` of a blank stays
blank.

`UPCASE`, `LOWCASE`, and `PROPCASE` follow the SAS locale and translation
table. The examples on this page are ASCII `A-Z`. `toupper()` and `tolower()`
follow the R locale for letters outside that range.
