---
title: "Transpose rows and columns"
subtitle: "pivot_longer() / pivot_wider()"
---

::: {.sas-equiv}
**SAS:** `PROC TRANSPOSE` (`BY`, `ID`, `VAR`, `NAME=`, `PREFIX=`, `LET`),
then `MERGE` / SQL `JOIN` of `SUPPxx` onto the parent domain
:::

`PROC TRANSPOSE` turns selected columns into rows. With `ID`, the values of
that variable become column names. This site uses `tidyr::pivot_longer()` and
`tidyr::pivot_wider()` for that pair. Base `reshape()` is not the form used
here.

## Quick reference

{{< include _quickref/transpose.md >}}

Worked examples below.

## Long to wide

One row per subject and lab test. The wide table has one row per subject and
one column per `PARAMCD`.

```{r}
#| label: setup
library(dplyr)
library(tidyr)

lb <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0001", "STD-001-0002", "STD-001-0002"),
  PARAMCD = c("ALT", "AST", "ALT", "AST"),
  AVAL = c(22, 24, 31, 28)
)

wide <- lb |>
  pivot_wider(names_from = PARAMCD, values_from = AVAL)

wide
```

SAS:

```sas
proc sort data=lb;
  by usubjid;
run;

proc transpose data=lb out=lb_wide(drop=_name_);
  by usubjid;
  id paramcd;
  var aval;
run;
```

`BY` variables must be sorted before `PROC TRANSPOSE`. `ID` values become the
column names `ALT` and `AST`. `VAR` supplies the cells. `_NAME_` holds the
source variable name (`AVAL`); the `drop=` removes it. `pivot_wider()` does not
add `_NAME_` and does not require a sort.

`PREFIX=AVAL_` together with `ID` names the columns `AVAL_ALT` and `AVAL_AST`.
In R that prefix is `names_prefix = "AVAL_"`.

## Wide to long

```{r}
#| label: longer
wide |>
  pivot_longer(c(ALT, AST), names_to = "PARAMCD", values_to = "AVAL")
```

SAS:

```sas
proc transpose data=lb_wide out=lb_long(rename=(col1=aval)) name=paramcd;
  by usubjid;
  var alt ast;
run;
```

`NAME=` is `names_to`. Without an `ID` statement the values sit in `COL1`.
Renaming `COL1` is `values_to`. Without `ID`, `PREFIX=` replaces `COL` in
`COL1`, `COL2`, .... That is a different naming rule from `PREFIX=` used with
`ID`.

## Duplicate `ID` values

`STD-001-0001` now has two `ALT` rows.

```{r}
#| label: dup
lb_dup <- rbind(
  lb,
  data.frame(USUBJID = "STD-001-0001", PARAMCD = "ALT", AVAL = 99)
)

lb_dup |>
  pivot_wider(names_from = PARAMCD, values_from = AVAL)
```

`ALT` for `STD-001-0001` is a list of two numbers, not one result.
`PROC TRANSPOSE` without `LET` stops when an `ID` value occurs twice in the
same `BY` group. `LET` keeps the last occurrence in that group.

```{r}
#| label: let
lb_dup |>
  pivot_wider(names_from = PARAMCD, values_from = AVAL, values_fn = last)
```

`values_fn = last` keeps 99, the last `ALT` for `STD-001-0001`. That is the
`LET` result. It does not check which row is the right one.

## `SUPPAE` onto `AE`

`SUPPxx` is SDTM. Each row is one supplemental qualifier as collected: `QNAM`
is the name, `QVAL` is the value. `QVAL` is character, including when the text
looks numeric. `IDVAR` names the parent key (`AESEQ`). `IDVARVAL` is that key's
value, stored as character. A blank `IDVAR` means the qualifier is
subject-level, and the key is `STUDYID` plus `USUBJID` only.

This is not an ADaM imputation. `AETRTEM` and `AENDAY` are the qualifier names
in this example. `AENDAY` is sponsor-defined: a day count, stored as text. It
is not an AE domain variable. A standard AE variable such as `AEDUR` stays on
AE. The IG does not allow it in SUPPQUAL, and `AEDUR` is an ISO 8601 duration,
not a number of days.

```{r}
#| label: supp-data
ae <- data.frame(
  STUDYID = "STD-001",
  USUBJID = c("STD-001-0001", "STD-001-0001", "STD-001-0002"),
  AESEQ = c(1L, 2L, 1L),
  AETERM = c("Headache", "Nausea", "Fatigue"),
  AESEV = c("MILD", "MODERATE", "MILD")
)

suppae <- data.frame(
  STUDYID = "STD-001",
  RDOMAIN = "AE",
  USUBJID = c(
    "STD-001-0001", "STD-001-0001", "STD-001-0001",
    "STD-001-0001", "STD-001-0002"
  ),
  IDVAR = "AESEQ",
  IDVARVAL = c("1", "1", "2", "2", "1"),
  QNAM = c("AETRTEM", "AENDAY", "AETRTEM", "AENDAY", "AETRTEM"),
  QVAL = c("Y", "5", "Y", "", "N")
)

ae
suppae
```

`STD-001-0001` sequence 2 has an `AENDAY` row whose `QVAL` is blank.
`STD-001-0002` sequence 1 has no `AENDAY` row.

Keep record-level qualifiers, convert `IDVARVAL` to the parent type, then
widen. See [Convert character, numeric, and date](type-convert.qmd) for
`as.integer()` and for turning a numeric-looking `QVAL` into a number
afterwards.

```{r}
#| label: supp-wide
supp_wide <- suppae |>
  filter(RDOMAIN == "AE", IDVAR == "AESEQ") |>
  mutate(AESEQ = as.integer(IDVARVAL)) |>
  pivot_wider(
    id_cols = c(STUDYID, USUBJID, AESEQ),
    names_from = QNAM,
    values_from = QVAL
  )

supp_wide
```

`AETRTEM` and `AENDAY` are character. The blank `AENDAY` is `""`. The absent
`AENDAY` is `NA`.

SAS, after sorting by the `BY` variables:

```sas
proc transpose data=suppae(where=(rdomain="AE" and idvar="AESEQ"))
               out=supp_wide(drop=_name_);
  by studyid usubjid idvarval;
  id qnam;
  var qval;
run;

data supp_keyed;
  set supp_wide;
  aeseq = input(idvarval, best.);
run;
```

`IDVARVAL` is character. `AESEQ` on `AE` is numeric. `input(idvarval, best.)`
is the SAS conversion before a `MERGE` on `studyid`, `usubjid`, and `aeseq`.
Sort both tables by those keys first.

Joining `AESEQ` (integer) to `IDVARVAL` (character) fails.

```{r}
#| label: bad-join
#| error: true
supp_bad <- suppae |>
  filter(RDOMAIN == "AE", IDVAR == "AESEQ") |>
  pivot_wider(
    id_cols = c(STUDYID, USUBJID, IDVARVAL),
    names_from = QNAM,
    values_from = QVAL
  )

left_join(
  ae,
  supp_bad,
  by = c("STUDYID", "USUBJID", "AESEQ" = "IDVARVAL"),
  relationship = "one-to-one"
)
```

Convert first, then join on `STUDYID`, `USUBJID`, and `AESEQ`. The join rules
are [Merge on keys](joins.qmd).

```{r}
#| label: join
ae_plus <- left_join(
  ae,
  supp_wide,
  by = c("STUDYID", "USUBJID", "AESEQ"),
  relationship = "one-to-one"
)

ae_plus
```

::: {.callout-important}
## A blank `QVAL` is not a missing SUPP row

Sequence 2 for `STD-001-0001` has `AENDAY` equal to `""`: the qualifier was
submitted blank. Sequence 1 for `STD-001-0002` has `AENDAY` equal to `NA`:
`SUPPAE` has no `AENDAY` row for that event. `pivot_wider()` does not turn `""`
into `NA`. In SAS both results are character missing, so the transposed column
cannot show the difference.
:::

`QVAL` is still character after the join. A sponsor-defined day count has to
be converted before it is used as a number. `as.numeric(na_if(AENDAY, ""))`
turns the blank into `NA`, and the already missing qualifier stays `NA`. After
that step the two reasons are gone. Keep the character value, or a flag, if a
later step must tell them apart.

```{r}
#| label: aenday
ae_plus |>
  mutate(AENDAY = as.numeric(na_if(AENDAY, "")))
```
