---
title: "Convert character, numeric, and date"
subtitle: "as.character() / as.numeric() / as.Date()"
---

::: {.sas-equiv}
**SAS:** `PUT`, `INPUT`, informats and formats (`BEST.`, `Zw.`, `YYMMDD10.`, `E8601DT.`)
:::

`PUT` writes a value with a format and always returns character. `INPUT` reads a
character value with an informat. The informat sets the result type. R uses one
function per target type.

## Quick reference

{{< include _quickref/type-convert.md >}}

Worked examples below.

## Character and numeric

```{r}
#| label: setup
library(dplyr)

raw <- data.frame(
  USUBJID = c("STD-001-0001", "STD-001-0002", "STD-001-0003", "STD-001-0004"),
  SITEC = c("001", "010", "002", "003"),
  AGEC = c("54", "61", "", " "),
  RESC = c("1.25", "1,25", "12", "8")
)

raw
```

`SITEC` is character, including leading zeros. `AGEC` has a zero-length string
and a single space. `RESC` has one decimal point and one decimal comma.

```{r}
#| label: site-age
raw |>
  mutate(
    SITE = as.numeric(SITEC),
    AGE = as.numeric(na_if(trimws(AGEC), ""))
  )
```

SAS:

```sas
site = input(sitec, best.);
age = input(agec, best.);
```

`SITE` is 1, 10, 2, 3. `"001"` and `"010"` lose their leading zeros because a
numeric has no zeros to keep. `AGE` is 54, 61, then `NA` for both the empty
string and the space. `trimws()` makes an all-blank string `""`, and `na_if()`
makes that `NA` before `as.numeric()`, so this path does not warn.

Passed straight to `as.numeric()`, those blanks still become `NA`. There is no
warning. A blank is not the same thing as text that cannot be a number.

```{r}
#| label: blank-num
as.numeric(c("54", "", " "))
```

```{r}
#| label: as-character
as.character(1)
```

`as.character(1)` is `"1"`. SAS `strip(put(1, best.))` is `"1"`. `put(1, best.)`
without `strip` is `BEST12.`, so the result can contain leading blanks.
`as.character()` does not pad.

::: {.callout-important}
## Blank is not `NA`

SAS character missing is blank. `missing()` is true when a character value is
all blanks. In R, `""` and `" "` are present. Only `NA` is missing.
`as.numeric("")` returns `NA` and does not warn. SAS `input("", best.)` also
returns numeric missing without an invalid-data note. The character value was
not `NA` before the call.
:::

## Decimal comma

```{r}
#| label: comma
raw |>
  mutate(RES = as.numeric(RESC))
```

`1.25`, `NA`, `12`, `8`. `"1,25"` is not a number to `as.numeric()`. `mutate()`
warns `NAs introduced by coercion` for that value. A blank does not warn;
`"1,25"` does. SAS `input("1,25", best.)` also returns missing, and it writes
an invalid-data note. `input("1,25", ?? best.)` returns missing and suppresses
the note. Neither reads the comma as a decimal mark.

SAS `input("1,25", commax5.)` reads 1.25. `COMMAX` uses a comma as the decimal
mark and a period as the thousands separator. `COMMA` is the opposite.

```{r}
#| label: comma-fix
as.numeric(sub(",", ".", "1,25", fixed = TRUE))
```

`sub()` here replaces one comma. Use it only when the comma is the decimal mark
and there is no thousands separator. It is not a general `COMMAX` parser.
`as.numeric()` does not follow `options(OutDec = ",")`. That option changes how
numbers are printed, not how `as.numeric()` reads them.

## Put the zeros back

The numeric value cannot store the width. Write it on the way out, as `Zw.` does.

```{r}
#| label: zeros
raw |>
  mutate(
    SITE = as.integer(SITEC),
    SITE_CHAR = as.character(SITE),
    SITEC2 = sprintf("%03d", SITE)
  )
```

`as.integer("001")` is `1`. `as.character()` of that value is `"1"`.
`sprintf("%03d", SITE)` is `"001"`. SAS `put(site, z3.)` writes the same
character zeros.

`%03d` needs an integer. A non-integer numeric errors. Missing becomes
`" NA"`, not a blank:

```{r}
#| label: sprintf-na
sprintf("%03d", c(1L, NA_integer_))
```

## Factors

`as.numeric()` on a factor uses the level codes.

```{r}
#| label: factor
fac <- factor(c("10", "20"))
as.numeric(fac)
as.numeric(as.character(fac))
```

The codes are 1 and 2. The labels are `"10"` and `"20"`. Convert the labels:
`as.numeric(as.character(fac))`.

## Dates

A complete ISO date string and a SAS date number are different types. Partial
`--DTC` strings are a separate problem: see [Partial dates](partial-dates.qmd).
Study day on `Date` values is [Study day](study-day.qmd).

```{r}
#| label: iso-date
as.Date(c("2025-06-03", "2025-06-10", ""))
```

`as.Date("2025-06-03")` is that calendar day. `as.Date("")` is `NA`.
SAS `input("2025-06-03", yymmdd10.)` does not return a `Date`. It returns the
integer day count from 1960-01-01. `as.Date()` is that `INPUT` direction only.

```{r}
#| label: date-put
as.character(as.Date("2025-06-03"))
format(as.Date("2025-06-03"), "%Y-%m-%d")
```

Both are the character `2025-06-03`. That is `put(date, yymmdd10.)`.

```{r}
#| label: sas-date
sas_n <- as.integer(as.Date("2025-06-03") - as.Date("1960-01-01"))
sas_n
as.Date(sas_n, origin = "1960-01-01")
as.Date(sas_n)
```

`sas_n` is 23895. `origin = "1960-01-01"` returns 2025-06-03. With no `origin`,
`as.Date()` counts from 1970-01-01 and returns 2035-06-04.

::: {.callout-important}
## A SAS date number is not an R Date

`as.Date(n)` with no `origin` means days since 1970-01-01. A SAS date is days
since 1960-01-01. The same integer is not the same calendar day. Pass
`origin = "1960-01-01"` when `n` came from SAS.
:::

The other direction, from an R `Date` to a SAS date number:

```{r}
#| label: to-sas-date
as.integer(as.Date("2025-06-03") - as.Date("1960-01-01"))
```

## Datetime

SAS datetime values are seconds since 1960-01-01 00:00:00.
`input("2025-06-03T14:30:00", e8601dt.)` reads that ISO form.

```{r}
#| label: posix
txt <- "2025-06-03T14:30:00"
as.POSIXct(txt, tz = "UTC")
as.POSIXct(txt, format = "%Y-%m-%dT%H:%M:%S", tz = "UTC")

sas_dt <- as.numeric(
  as.POSIXct("2025-06-03 14:30:00", tz = "UTC") -
    as.POSIXct("1960-01-01", tz = "UTC"),
  units = "secs"
)
as.POSIXct(sas_dt, origin = "1960-01-01", tz = "UTC")
```

With no `format`, `as.POSIXct(txt, tz = "UTC")` does not error. It returns
`2025-06-03 UTC`. The date parses and the time is dropped. Pass
`format = "%Y-%m-%dT%H:%M:%S"` to keep `14:30:00`.

Set `tz`. A `POSIXct` value is an instant. Another zone prints another clock
time. `"UTC"` keeps the clock time stored in a zone-naive SAS datetime.
